---
title: "covSTATIS vignette"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
rm(list=ls())
library(tidyr)
library(dplyr)
library(osfr)
library(readr)
library(DistatisR)  # covSTATIS analysis
library(PTCA4CATA)  # factor map plotting
library(ExPosition) # makeDataNominal()
library(gplots)     # heatmap.2()
library(sp)         # convex hulls 
library(ggplot2)
library(RColorBrewer)
library(superheat)
```

## Setup

All data used in this tutorial are stored at: https://github.com/giuliabaracc/covSTATIS_netneuro/tree/main/Data. The "Data" repo contains: 144x3 functional connectivity text files (n=144, 3=task conditions, 0-,1-,2-back), a "participants_ages.csv" file with participants info and an atlas labels file "Schaefer_100parcel_17network_labels.csv". 

## Data info

After standard preprocessing (see Rieck et al., 2021 Data in Brief), each individual’s voxel-time series were parcellated, for each task condition, using the standard 100 region-17 network Schaefer atlas (Schaefer et al., 2018 Cerebral Cortex). For each participant and condition, Pearson's correlations were quantified between region pairs, to ultimately obtain 100x100x3 functional connectivity matrices for each individual. 

## Read in the data

The first step is to read in our connectivity matrices to form a 3D array (= 'X' here below) for analysis with covSTATIS. In this tutorial, the 3D array dimensionality is 100 x 100 x 144 x 3 (ROI x ROI x N x Task condition). 

```{r}
setwd('../Data/')
atlas_prefix <- "Schaefer100_17"
atlas_labels <- read_csv('../Data/Schaefer_100parcel_17network_labels.csv')
atlas_labels$Network <- as.factor(atlas_labels$Network)
demog_in <- read_csv('../Data/participant_ages.csv')
ids <- demog_in$subjectID

## Read in FC .txt files and reshape into 3D array with size n_roi x n_roi x n_sub*n_conditions (i.e. n_fns)
all_fns<-list.files(path = "../Data/", pattern = paste0('^',atlas_prefix, '.*txt'))
X_list<-lapply(X=all_fns, FUN=read_delim, delim="\t", col_names = F,col_type = list(.default = col_double()))
X <- array(unlist(X_list), c(dim(X_list[[1]]), length(X_list))) #our 3D array
rm(X_list) 

## Label rows and columns with ROI label
if(length(grep("Schaefer", atlas_prefix)>0)){
  rownames(X)<- atlas_labels$ROI_label
  colnames(X)<- atlas_labels$ROI_label
}

## Label 3rd dimension with subject and task information (modified from filenames)
dimnames(X)[[3]]<-all_fns %>% gsub(pattern=".txt", replacement = "") %>%
  gsub(pattern=paste0(atlas_prefix,"_"), replacement = "")

```

## Run covSTATIS

Now that we have created our 3D data array (X), we can run covSTATIS. Since we are dealing with correlation values, the Distance flag should be set to FALSE. The nfact2keep flag denotes the number of components we want to keep (the bigger the value, the slower the computations).

```{r}
covstatis_res <- distatis(X, Distance = F,nfact2keep = 10,compact = F)
```

### Step 1: RV similarity matrix 

The first operation covSTATIS computes is the generation of an RV similarity matrix. The RV matrix quantifies the pairwise similarity among all data tables (squared Pearson's correlation), irrespective of their rotation or scaling. The RV matrix then undergoes eigenvalue decomposition (EVD). The resulting first component best represents the common pattern across all tables, and its first eigenvector quantifies how similar each table is to this common pattern. 

Let's visualize the results of this EVD on the RV matrix. The Scree plot shows how much variance is explained by each component/dimension. Ideally, there will be one strong component (as is the case with these data) indicating a strong coherence in the general network organization across all observations.

```{r}
PlotScree(ev = covstatis_res$res4Cmat$eigValues, 
          title = "RV-map: Explained Variance per Dimension")
```

The factor map of the RV space shows how similar the data tables are to each other.

```{r}
rv.labels <- createxyLabels.gen(x_axis = 1,
                                y_axis = 2,
                                lambda = covstatis_res$res4Cmat$eigValues,
                                tau = covstatis_res$res4Cmat$tau)
## factor map
rv.map <- createFactorMap(covstatis_res$res4Cmat$G,
                          title = "The component space of the RV matrix")

rv.map$zeMap_background + rv.map$zeMap_dots + rv.labels

```

### Step 2: Compromise space

Next, covSTATIS grabs the first eigenvector of this EVD on RV and scales it to sum to 1. The compromise matrix is built from the weighted sum of all data matrices and then also submitted to EVD. Let's first visualize the compromise space/matrix.

```{r}
net.col <- as.data.frame(unique(atlas_labels[,c(3,4)]))
rownames(net.col) <- net.col$Network

col.pal <- brewer.pal(11, "RdBu")
superheat(covstatis_res$res4Splus$Splus,
          membership.cols = atlas_labels$Network,
          membership.rows = atlas_labels$Network,
          clustering.method = NULL,
          heat.lim = c(-0.1,0.1),
          heat.pal = rev(brewer.pal(11, "RdBu")),
          heat.pal.values = c(0,0.45,0.5,0.55,1),
          left.label.size = 0.08,
          bottom.label.size = 0.05,
          y.axis.reverse = TRUE,
          left.label.col = net.col[levels(atlas_labels$Network),"Network_color"], # order by community name
          bottom.label.col =  net.col[levels(atlas_labels$Network), "Network_color"],
          left.label.text.size = 3,
          bottom.label.text.size = 2,
          left.label.text.col = "black",
          bottom.label.text.col = "black",
          left.label.text.alignment = "left",
          title = "The compromise")$plot
```

Next, the compromise undergoes EVD. Let's visualize its scree plot:

```{r}
# ### GB comment: this doesn't work. There is no info stored in covstatis_res$res4Splus$eigValues?
# PlotScree(ev = covstatis_res$res4Splus$eigValues, 
#           title = "Compromise: Explained Variance per Dimension")
```

Let's now reconstruct a heatmap to explore the dimensions from the EVD on the compromise. Let's start by visualizing the first two components (LV1, LV2).

```{r}
covstatis_corrmat_lv1_lv2 <- tcrossprod(covstatis_res$res4Splus$F[,c(1,2)])

superheat(covstatis_corrmat_lv1_lv2,
          membership.cols = atlas_labels$Network,
          membership.rows = atlas_labels$Network,
          clustering.method = NULL,
          heat.lim = c(-0.1,0.1),
          heat.pal = rev(brewer.pal(11, "RdBu")),
          heat.pal.values = c(0,0.45,0.5,0.55,1),
          left.label.size = 0.08,
          bottom.label.size = 0.05,
          y.axis.reverse = TRUE,
          left.label.col = net.col[levels(atlas_labels$Network),"Network_color"], # order by community name
          bottom.label.col =  net.col[levels(atlas_labels$Network), "Network_color"],
          left.label.text.size = 3,
          bottom.label.text.size = 2,
          left.label.text.col = "black",
          bottom.label.text.col = "black",
          left.label.text.alignment = "left",
          title = "Rebuilt Heatmap for LV1 and LV2")$plot
```

### Step 3: global factor scores from compromise space

The compromise space shows us how regional connectivity values relate to each other across all data tables (individuals and conditions). Each point in the compromise space represents a global factor score: a different ROI from our functional connectivity matrices, colored by their network membership. The closer points are, the stronger the functional connectivity across the entire dataset.

```{r}
network_labels <- unique(atlas_labels$Network)
network_colors <- unique(atlas_labels$Network_color)

# To get graphs with axes 1 and 2:
h_axis <-1 # component to plot on x-axis
v_axis <-2 # component to plot on y-axis

compromise_graph_out <- createFactorMap(covstatis_res$res4Splus$F,
                                        axis1 = h_axis, axis2 = v_axis,
                                        title = 'Compromise ROI Space',
                                        col.points = atlas_labels$Network_color,
                                        alpha.points=.6, cex=3) 
print(compromise_graph_out$zeMap_background + compromise_graph_out$zeMap_dots)
```

We can also compute the average of all the ROIs (i.e., barycenter) for each _a priori_ network to get a coarser network compromise map.

```{r}
network_mean_F <- t(apply(makeNominalData(as.matrix(atlas_labels$Network)),2,function(x){x/sum(x)})) %*% covstatis_res$res4Splus$F
row.names(network_mean_F) <- network_labels

compromise_network_graph_out <- createFactorMap(network_mean_F,
                                                axis1 = h_axis,axis2 = v_axis, 
                                                title = "Compromise map for network-level barycenters",
                                                col.points = network_colors, col.labels = network_colors,
                                                alpha.points=.8, cex=5, text.cex=4, alpha.labels=.8, pch=18)
print(compromise_network_graph_out$zeMap + compromise_network_graph_out$zeMap_dots)
```

Both the ROI-level and network-level factor maps can be overlaid to show how individuals ROIs cluster around their respective networks, for the whole sample.

```{r}
print(compromise_graph_out$zeMap_background + compromise_graph_out$zeMap_dots + 
        compromise_network_graph_out$zeMap_dots + compromise_network_graph_out$zeMap_text)
```

### Step 4: partial factor scores from compromise space

Using partial projection, we can now project each condition back onto the compromise space (i.e., partial factor scores) to understand how each correlation matrix fits in the compromise (or sample-level) space. 

To do that, we should first create a design matrix specifying which slice in our X cube corresponds to which n-back condition:

```{r}
condition_design <- vector("character",dim(X)[3])
condition_design[grep("*_nbk_0b",dimnames(X)[[3]])]<-"0b"
condition_design[grep("*_nbk_1b",dimnames(X)[[3]])]<-"1b"
condition_design[grep("*_nbk_2b",dimnames(X)[[3]])]<-"2b"
```

Now we can compute and plot the partial maps for the different n-back conditions, for each ROI. This plot shows how much the connectivity for each ROI is similar/different across task conditions (closer to the ROI center: more similar; farther from ROI center: more different). 

```{r}
#Compute the Partial map
F_j     <- covstatis_res$res4Splus$PartialF
alpha_j <- covstatis_res$res4Cmat$alpha
# create the groups based on design
code4Groups <- unique(condition_design)

nK <- length(code4Groups)
# initialize F_K and alpha_k
F_k <- array(0, dim = c(dim(F_j)[[1]], dim(F_j)[[2]],nK))
dimnames(F_k) <- list(dimnames(F_j)[[1]], 
                      dimnames(F_j)[[2]], code4Groups)
alpha_k <- rep(0, nK)
names(alpha_k) <- code4Groups
Fa_j <- F_j
for (j in 1:dim(F_j)[[3]]){ Fa_j[,,j]  <- F_j[,,j] * alpha_j[j] }
for (k in 1:nK){
  lindex <- condition_design == code4Groups[k]
  alpha_k[k] <- sum(alpha_j[lindex])
  F_k[,,k] <- (1/alpha_k[k])*apply(Fa_j[,,lindex],c(1,2),sum)
}

compromise_pfs <- createPartialFactorScoresMap(
  factorScores = covstatis_res$res4Splus$F,      
  partialFactorScores = F_k,  
  axis1 = 1, axis2 = 2,
  colors4Items = as.vector(atlas_labels$Network_color), 
  colors4Blocks = c("yellowgreen", "olivedrab", "darkgreen"),
  names4Partial = dimnames(F_k)[[3]], # 
  shape.points = 20)

print(compromise_graph_out$zeMap_background + compromise_graph_out$zeMap_dots +
  compromise_pfs$mapColByBlocks + ggtitle('Compromise map - Partial factor scores by task design'))
```

That's a very busy plot! Let's clean it up to make things easier to interpret. We can compute the partial factor scores by the mean network compromise. Similar to our ROI plot, for each network, the more spread out the conditions are from the network mean, the more diverse the functional connectivity is across task conditions. Tip: if you have groups, sometimes it is helpful to separate partial factor scores per group and plot them individually (e.g., you may not be able to see a difference if everything is plotted together).

```{r}
F_k_network<-array(NA, dim=c(length(network_labels), dim(F_k)[2], dim(F_k)[3]))
dimnames(F_k_network)<-list(network_labels, dimnames(F_k)[[2]], dimnames(F_k)[[3]])

for(i in 1:dim(F_k_network)[3]){
  F_k_network[,,i]<- t(apply(makeNominalData(as.matrix(atlas_labels$Network)),2,function(x){x/sum(x)})) %*% F_k[,,i]
}

compromise_pfs_network <- createPartialFactorScoresMap(
  factorScores = network_mean_F,      
  partialFactorScores = F_k_network,  
  axis1 = 1, axis2 = 2,
  colors4Items = network_colors, 
  colors4Blocks = c("yellowgreen", "olivedrab", "darkgreen"),
  names4Partial = dimnames(F_k_network)[[3]], # 
  size.labels=3.5, size.points = 2,shape.points=20)

print(compromise_network_graph_out$zeMap + compromise_pfs_network$mapColByBlocks + ggtitle('Network level compromise map - Partial factor scores by task design'))
```

### Step 5: area of the convex hull

We can now compute the area of the convex hull around the partial factor scores for each individual at the network-level (NB: this can be done at the regional level too). Area of the hull scores can be seen as "task condition differentiation scores": the larger the area of the hull, the more diverse the network functional connectivity is across task conditions (more distant from mean); the smaller the area of the hull, the more homogenous the functional connectivity across 0-,1-, and 2-back. Note: area of the hull values are arbitrary so you can go ahead and normalize them. Make sure you have at least 4 points to build the area of the hull (here you'll see warnings cause some networks have less than 4 observations/ROIs).

```{r}
## Create empty matrix to put results in, should be # participants (rows) by # networks (columns)
task_diff_network_level <-data.frame(matrix(0, length(ids),length(network_labels)))
rownames(task_diff_network_level) <- ids
names(task_diff_network_level) <- network_labels

partialF <- covstatis_res$res4Splu$PartialF

for(i in 1:length(ids)){
  
  ### Grab all the partialF values associated with this subject for LV1
  this_lv1 <- partialF[,1,grep(paste0(ids[i], "_*"),dimnames(partialF)[[3]])]
  ## Transform the node-level partialF to network-level partialF for LV1
  this_lv1_network<-t(apply(makeNominalData(as.matrix(atlas_labels$Network)),2,function(x){x/sum(x)})) %*% 
    as.matrix(this_lv1)
  
  ### Grab all the partialF values associated with this P for LV2
  this_lv2 <- partialF[,2,grep(paste0(ids[i], "_*"),dimnames(partialF)[[3]])]
  ## Transform the node-level partialF to network-level partialF for LV2
  this_lv2_network<-t(apply(makeNominalData(as.matrix(atlas_labels$Network)),2,function(x){x/sum(x)})) %*% 
    as.matrix(this_lv2)
  
  ## Loop through each network, get the LV1 and LV2 points, and compute the area of the hull around the points
  for(n in 1:length(network_labels)){
    hull_pts <- chull(this_lv1_network[n,], this_lv2_network[n,]) #here you can play with the flag "percentage" and include either all points (like done here) or e.g., 95% of them in which case you'd write "percentage=0.95"
    this_hull_coords <- cbind(this_lv1_network[n,hull_pts],this_lv2_network[n,hull_pts] )
    this_poly_area <- Polygon(this_hull_coords, hole=F)@area
    task_diff_network_level[i,n] <- this_poly_area
    
  }
} 

#print(task_diff_network_level) #you can go ahead and save this output and relate it to e.g., behavior
```

Area of the hull scores can be used in secondary analyses, such as brain-behavior correlation. Since our sample is an adult lifespan sample, we can relate area of the hull scores to age. This plot shows how Default A functional connectivity becomes more homogeneous/less differentiated across tasks as a function of age.

```{r}
secondary_analyses <- cbind(demog_in$Age, task_diff_network_level)
secondary_analyses <- as.data.frame(secondary_analyses)
colnames(secondary_analyses)[1] <- "Age"
colnames(secondary_analyses)[14] <- "DefaultA"
ggplot(secondary_analyses, aes(x=Age, y=DefaultA)) + 
  geom_point()+
  geom_smooth(method=lm)
```